---
title: "Egg neglect behaviour in biparental seabird, the Little Auk *Alle alle*"
author: "Martyan Gajzmer"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: lumen
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: hide
    number_sections: true
---

# Diurnal pattern of exits (neglect, frequency) in Little Auk

In this section, we analyze the pattern of exiting the nest during incubation to determine the general rhythm and if there are sex-specific differences in the diurnal rhythm of nest exits.

## Model and results

To model the diurnal pattern of exits, we used a Bayesian Poisson GLM with circular (sine/cosine) encoding of hour, which correctly handles the cyclic nature of time-of-day (i.e., hour 23 is adjacent to hour 0). The model structure is defined as follows:

$$\begin{aligned}
G_{i} &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \alpha_{sex[i]} + \beta^{sin}_{sex[i]} \sin(h_i) + \beta^{cos}_{sex[i]} \cos(h_i) \\
h_i &= \frac{2\pi \cdot \text{hour}_i}{24} \\
\alpha_j &\sim \text{Normal}(0, 1) \\
\beta^{sin}_j &\sim \text{Normal}(0, 1) \\
\beta^{cos}_j &\sim \text{Normal}(0, 1)
\end{aligned}$$

Where $G_i$ is the number of gaps (exits) per hour, $\alpha_j$ is the sex-specific intercept, and $\beta^{sin}_j$, $\beta^{cos}_j$ are sex-specific weights on the sine and cosine projections of hour, capturing the amplitude and phase of the daily cycle.

```{r load-libraries, message=FALSE, warning=FALSE}
library(rethinking)
library(tidyverse)
library(lubridate)
library(splines)
library(dplyr)

# Global Visual Settings
stage_names  <- c("early", "mid", "late")
stage_colors <- c("#3B6D11", "#185FA5", "#993556")
sex_labels   <- c("Female", "Male")
sex_col      <- c("darkorange", "navyblue")

# ------------------------------------------------------------------------------
# NEST-LEVEL DATA PREPARATION
# ------------------------------------------------------------------------------
E <- readRDS("./EDA_exit_hourly.RDS")

E <- E %>%
  mutate(
    incubation_stage = case_when(
      day_prior_hatch >= 21 & day_prior_hatch <= 30 ~ "early",
      day_prior_hatch >= 11 & day_prior_hatch <= 20 ~ "mid",
      day_prior_hatch >= 0  & day_prior_hatch <= 10 ~ "late",
      TRUE ~ NA_character_
    ),
    stage_id = as.integer(factor(incubation_stage, levels = stage_names)),
    pair_id  = as.integer(factor(Pair_status, levels = c("New", "Old"))),
    stage_pair_id = as.integer(factor(
      paste(incubation_stage, Pair_status, sep = "_"),
      levels = c("early_New", "early_Old", "mid_New", "mid_Old", "late_New", "late_Old")
    ))
  )

# ------------------------------------------------------------------------------
# SEX-SPECIFIC DATA PREPARATION
# ------------------------------------------------------------------------------
S <- readRDS("EDA_sex_neglect_wide.RDS")
G <- readRDS("EDA_sex_gaps_hourly.RDS")

sex_hourly_raw <- G %>%
  filter(!is.na(responsible_sex)) %>%
  group_by(season, session, nest, sx = responsible_sex, hour) %>%
  summarise(n_gaps_per_hour = n(), .groups = "drop")

meta <- S %>%
  mutate(
    season = as.character(season),
    incubation_stage = case_when(
      day_prior_hatch >= 21 & day_prior_hatch <= 30 ~ "early",
      day_prior_hatch >= 11 & day_prior_hatch <= 20 ~ "mid",
      day_prior_hatch >= 0  & day_prior_hatch <= 10 ~ "late",
      TRUE ~ NA_character_
    ),
    stage_id = as.integer(factor(incubation_stage, levels = stage_names)),
    sex_id   = as.integer(factor(sx, levels = c("f", "m"))),
    stage_sex_id = as.integer(factor(paste(incubation_stage, sx, sep = "_"),
                                     levels = c("early_f", "early_m", "mid_f", "mid_m", "late_f", "late_m")))
  ) %>%
  select(season, session, nest, sx, day_prior_hatch, incubation_stage, stage_id, sex_id, stage_sex_id)

E_sex <- meta %>% crossing(hour = 0:23) %>%
  left_join(sex_hourly_raw, by = c("season", "session", "nest", "sx", "hour")) %>%
  mutate(n_gaps_per_hour = replace_na(n_gaps_per_hour, 0))

# Add circular encoding to model data
E_fit_sex <- E_sex %>%
  filter(!is.na(sex_id)) %>%
  mutate(
    hour_rad = 2 * pi * hour / 24,
    sin_hour = sin(hour_rad),
    cos_hour = cos(hour_rad)
  )

# Prediction grid: 24 integer hours for summaries, 200 points for plotting
pred_grid_plot <- data.frame(hour = seq(0, 23, length.out = 200)) %>%
  mutate(
    hour_rad = 2 * pi * hour / 24,
    sin_hour = sin(hour_rad),
    cos_hour = cos(hour_rad)
  )

pred_grid_summary <- data.frame(hour = 0:23) %>%
  mutate(
    hour_rad = 2 * pi * hour / 24,
    sin_hour = sin(hour_rad),
    cos_hour = cos(hour_rad)
  )
```

```{r fit-m-circ, include=FALSE, cache=TRUE}
d_circ <- list(
  n_gaps_per_hour = E_fit_sex$n_gaps_per_hour,
  sex_id   = E_fit_sex$sex_id,
  sin_hour = E_fit_sex$sin_hour,
  cos_hour = E_fit_sex$cos_hour
)

m_circ <- quap(
  alist(
    n_gaps_per_hour ~ dpois(lambda),
    log(lambda) <- a[sex_id] +
                   b_sin[sex_id] * sin_hour +
                   b_cos[sex_id] * cos_hour,
    a[sex_id]     ~ dnorm(0, 1),
    b_sin[sex_id] ~ dnorm(0, 1),
    b_cos[sex_id] ~ dnorm(0, 1)
  ),
  data  = d_circ,
  start = list(a = rep(0,2), b_sin = rep(0,2), b_cos = rep(0,2))
)
```

```{r posterior-predictions, include=TRUE}
# Posterior for plotting (200-point grid)
post_f_plot <- link(m_circ, data = c(list(sex_id = rep(1, nrow(pred_grid_plot))),
                                      as.list(pred_grid_plot[, c("sin_hour","cos_hour")])))
post_m_plot <- link(m_circ, data = c(list(sex_id = rep(2, nrow(pred_grid_plot))),
                                      as.list(pred_grid_plot[, c("sin_hour","cos_hour")])))

pred_grid_plot$f_mu   <- apply(post_f_plot, 2, mean)
pred_grid_plot$f_low  <- apply(post_f_plot, 2, PI, prob = 0.89)[1,]
pred_grid_plot$f_high <- apply(post_f_plot, 2, PI, prob = 0.89)[2,]
pred_grid_plot$m_mu   <- apply(post_m_plot, 2, mean)
pred_grid_plot$m_low  <- apply(post_m_plot, 2, PI, prob = 0.89)[1,]
pred_grid_plot$m_high <- apply(post_m_plot, 2, PI, prob = 0.89)[2,]

# Posterior for numerical summaries (24-hour grid)
post_f_sum <- link(m_circ, data = c(list(sex_id = rep(1, 24)),
                                     as.list(pred_grid_summary[, c("sin_hour","cos_hour")])))
post_m_sum <- link(m_circ, data = c(list(sex_id = rep(2, 24)),
                                     as.list(pred_grid_summary[, c("sin_hour","cos_hour")])))

# Peak hour per posterior sample
for (s in 1:2) {
  post_s <- if (s == 1) post_f_sum else post_m_sum
  peak_h <- apply(post_s, 1, which.max) - 1
  cat(sex_labels[s], "Peak Hour:", round(mean(peak_h), 1), "\n")
}
```


```{r results-exit-pattern}
post_samples <- extract.samples(m_circ)

# Phase = atan2(b_sin, b_cos) converted to hours
phase_f <- atan2(post_samples$b_sin[,1], post_samples$b_cos[,1]) * 24 / (2 * pi)
phase_m <- atan2(post_samples$b_sin[,2], post_samples$b_cos[,2]) * 24 / (2 * pi)
phase_diff <- phase_f - phase_m

cat("Female phase (hours):", round(mean(phase_f), 2),
    " 89% PI [", round(PI(phase_f, prob = 0.89), 2), "]\n")
cat("Male phase (hours):  ", round(mean(phase_m), 2),
    " 89% PI [", round(PI(phase_m, prob = 0.89), 2), "]\n")
cat("Phase difference (F-M):", round(mean(phase_diff), 2),
    " 89% PI [", round(PI(phase_diff, prob = 0.89), 2), "]\n")
cat("P(|phase diff| < 1 hour):", round(mean(abs(phase_diff) < 1), 3), "\n")
```



Peak exit timing did not differ credibly between sexes (phase difference: 0.31 h, 89% PI [−1.66, 2.26]), whereas exit rate was consistently higher in males across all hours (difference: −0.034 gaps/hour, 89% PI [−0.043, −0.024]; P(Female < Male) = 1.0).


```{r results-mean-rates}
avg_rate_f_circ <- apply(post_f_sum, 1, mean)
avg_rate_m_circ <- apply(post_m_sum, 1, mean)
diff_fm_circ    <- avg_rate_f_circ - avg_rate_m_circ

cat("Female mean rate:", round(mean(avg_rate_f_circ), 3),
    " 89% PI [", round(PI(avg_rate_f_circ, prob = 0.89), 3), "]\n")
cat("Male mean rate:  ", round(mean(avg_rate_m_circ), 3),
    " 89% PI [", round(PI(avg_rate_m_circ, prob = 0.89), 3), "]\n")
cat("Difference (F-M):", round(mean(diff_fm_circ), 3),
    " 89% PI [", round(PI(diff_fm_circ, prob = 0.89), 3), "]\n")
cat("P(Female < Male):", round(mean(diff_fm_circ < 0), 3), "\n")
```

Males initiated gaps at a notably higher rate than females across the day (male mean: 0.121 gaps/hour, 89% PI [0.113, 0.128]; female mean: 0.087 gaps/hour, 89% PI [0.081, 0.093]). The difference was consistent and credible, with females showing reliably fewer exits than males (difference: −0.034, 89% PI [−0.043, −0.025]; P(Female < Male) = 1.0).

Both sexes showed a similar diel pattern, with gap frequency peaking in the early morning hours (female peak: ~5 AM, male peak: ~6 AM), suggesting a shared response to time-of-day cues rather than sex-specific scheduling of nest attendance.

```{r plot-hourly-gaps-by-sex, fig.width=7, fig.height=4, fig.cap="Expected number of exits per hour by sex"}
col_female <- "darkorange"
col_male   <- "navyblue"

obs_means <- E_fit_sex %>%
  group_by(sex_id, hour) %>%
  summarise(mean_gaps = mean(n_gaps_per_hour), .groups = "drop") %>%
  mutate(sex = ifelse(sex_id == 1, "f", "m"))

ggplot(pred_grid_plot, aes(x = hour)) +
  geom_ribbon(aes(ymin = f_low, ymax = f_high), fill = col_female, alpha = 0.15) +
  geom_ribbon(aes(ymin = m_low, ymax = m_high), fill = col_male,   alpha = 0.15) +
  geom_line(aes(y = f_mu), color = col_female, linewidth = 1.2) +
  geom_line(aes(y = m_mu), color = col_male,   linewidth = 1.2) +
  # Raw observed means as points
  geom_point(data = filter(obs_means, sex == "f"),
             aes(x = hour, y = mean_gaps), color = col_female, alpha = 0.4, size = 1.5) +
  geom_point(data = filter(obs_means, sex == "m"),
             aes(x = hour, y = mean_gaps), color = col_male,   alpha = 0.4, size = 1.5) +
  annotate("text", x = 0.5, y = max(pred_grid_plot$m_mu) * 1.13,
           label = "\u2642", size = 10, color = col_male) +
  annotate("text", x = 0.5, y = max(pred_grid_plot$f_mu) * 0.65,
           label = "\u2640", size = 10, color = col_female) +
  scale_x_continuous(breaks = seq(0, 23, by = 2), limits = c(0, 23)) +
  coord_cartesian(ylim = c(0, 0.25), clip = "off") +
  labs(x = "Hour of day", y = "Expected exits") +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Diurnal pattern of nest exits during incubation by sex. Lines show the posterior mean expected number of exits per hour for females (orange) and males (navy), derived from a Bayesian Poisson GLM with circular (sine/cosine) hour encoding. Shaded ribbons represent 89% posterior credible intervals. Points show observed hourly means averaged across all nest-sessions for each sex. Males exited the nest at a consistently higher rate than females across the day, while the overall shape of the diel cycle — including the early-morning peak — was similar between sexes.

```{r plot-hourly-gaps-by-sex2, fig.width=7, fig.height=4, fig.cap="Expected number of exits per hour by sex"}
ggplot(pred_grid_plot, aes(x = hour)) +
  geom_ribbon(aes(ymin = f_low, ymax = f_high), fill = col_female, alpha = 0.15) +
  geom_ribbon(aes(ymin = m_low, ymax = m_high), fill = col_male,   alpha = 0.15) +
  geom_line(aes(y = f_mu), color = col_female, linewidth = 1.2) +
  geom_line(aes(y = m_mu), color = col_male,   linewidth = 1.2) +
  geom_point(data = filter(obs_means, sex == "f"),
             aes(x = hour, y = mean_gaps), color = col_female, alpha = 0.4, size = 1.5) +
  geom_point(data = filter(obs_means, sex == "m"),
             aes(x = hour, y = mean_gaps), color = col_male,   alpha = 0.4, size = 1.5) +
  scale_x_continuous(breaks = seq(0, 23, by = 2), limits = c(0, 23)) +
  coord_polar() +
  labs(x = "Hour of day", y = "Expected exits") +
  theme_bw() +
  theme(panel.grid = element_blank())
```


# Neglect behaviour across incubation

In this section we model three complementary aspects of egg neglect across incubation days: how often gaps occur (frequency), whether any neglect happens at all on a given day (probability), and how long neglect lasts when it does occur (duration). Together these capture both the occurrence and intensity of neglect as hatching approaches.

## Gap frequency across incubation

We modelled the daily number of incubation gaps using a Poisson GLM with a linear effect of days prior to hatching, fit separately for each sex. Day was mean-centered prior to fitting.

$$\begin{aligned}
N_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \alpha_{sex[i]} + \beta_{sex[i]} \cdot \text{dph}_c \\
\alpha_j &\sim \text{Normal}(1,\, 0.5) \\
\beta_j  &\sim \text{Normal}(0,\, 0.05)
\end{aligned}$$

Where $N_i$ is the number of gaps on observation $i$, $\text{dph}_c$ is days prior to hatch (mean-centered), and parameters vary by sex $j$.

```{r prepare-neglect-data, include=FALSE}
S <- readRDS("EDA_sex_neglect_wide.RDS")
E <- readRDS("./EDA_egg_neglect_index_neglect_table.RDS")
G <- readRDS("./EDA_sex_gaps_table.RDS")


DAY_SEQ <- seq(2, 28, length.out = 100)

G_summary <- G %>%
  filter(dur >= 300) %>%
  group_by(season, session, nest, responsible_sex) %>%
  summarise(
    n_gaps = n(),
    sum_neglect_sec = sum(dur),
    mean_neglect_sec = mean(dur),
    .groups = "drop"
  ) %>%
  rename(sx = responsible_sex)

# Replace the neglect columns in S with the filtered versions
S_filt <- S %>%
  select(-n_gaps, -sum_neglect_sec, -mean_neglect_sec) %>%
  left_join(G_summary, by = c("season", "session", "nest", "sx")) %>%
  mutate(
    sum_neglect_sec = replace_na(sum_neglect_sec, 0),
    mean_neglect_sec = replace_na(mean_neglect_sec, 0),
    n_gaps = replace_na(n_gaps, 0L)
  )

E_fit_overall <- E %>%
  filter(!is.na(day_prior_hatch), !is.na(n_gaps),
         day_prior_hatch >= 2, day_prior_hatch <= 28) %>%
  mutate(dph_c = day_prior_hatch - mean(day_prior_hatch))

d_overall <- list(n_gaps = E_fit_overall$n_gaps, dph_c = E_fit_overall$dph_c)

E_fit_sex <- S_filt %>%
  filter(!is.na(day_prior_hatch), !is.na(sx),
         day_prior_hatch >= 2, day_prior_hatch <= 28) %>%
  mutate(
    sex_id = ifelse(sx == "f", 1, 2),
    dph_c  = day_prior_hatch - mean(day_prior_hatch)
  )

```

```{r duration_of_gaps}
#plot to see the duration of absences
G$duration_category <- cut(
  G$dur, 
  breaks = c(0, 300, 600, 900, 1800, 2700, 3600, 7200, Inf), 
  labels = c("0-5 min", "5-10 min", "10-15 min", "15-30 min", "30-45 min", "45-60 min", "1-2 hours", "2+ hours"),
  right = FALSE
)

# 1. Przygotowanie danych (upewnienie się, że kolory są zdefiniowane)
col_female <- "darkorange" 
col_male   <- "navyblue"

# 2. Tworzenie wykresu
ggplot(G[!is.na(G$duration_category), ], aes(x = duration_category, fill = responsible_sex)) +
  # Słupki z kolorami przypisanymi do płci
  geom_bar(color = "black", linewidth = 0.3, alpha = 0.8) +
  
  # Rozdzielenie na panele (płeć)
  facet_wrap(~responsible_sex, labeller = as_labeller(c(f = "Females", m = "Males"))) +
  
  # Manualne przypisanie kolorów, aby pasowały do linii trendu
  scale_fill_manual(values = c(f = col_female, m = col_male), guide = "none") +
  
  # Opisy
  labs(
    x = "Duration interval", 
    y = "Number of neglect gaps"
  ) +
  
  # --- THE "PRETTY" ADJUSTMENTS (Zgodne z p1, p2, p3) ---
  theme_bw(base_size = 12, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    
    # Formatowanie osi
    axis.title.x = element_text(size = 13, family = "sans", color = "black"),
    axis.title.y = element_text(size = 13, family = "sans", color = "black"),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, color = "black"),
    axis.text.y = element_text(size = 11, color = "black"),
    
    plot.title = element_text(face = "bold", size = 16),
    plot.margin = margin(10, 10, 10, 10)
  )

```


```{r prior-check-frequency, fig.width=6, fig.height=4, fig.cap="Prior predictive simulation for gap frequency model"}
set.seed(123)
dph_seq_c_ov <- DAY_SEQ - mean(E_fit_overall$day_prior_hatch)
n_sim <- 100
prior_mu_ov <- matrix(NA, nrow = n_sim, ncol = length(DAY_SEQ))


for(i in 1:n_sim){
  a <- rnorm(1, 1, 0.5)
  b <- rnorm(1, 0, 0.05)
  prior_mu_ov[i,] <- exp(a + b * dph_seq_c_ov)
}

matplot(DAY_SEQ, t(prior_mu_ov), type = "l", lty = 1, xlim = c(28, 2),
        col = col.alpha("black", 0.3),
        xlab = "Days prior hatch", ylab = "Expected n_gaps",
        main = "Prior check: gap frequency")#
```

```{r fit-frequency-models, include=FALSE, cache=TRUE}
d_overall <- list(n_gaps = E_fit_overall$n_gaps, dph_c = E_fit_overall$dph_c)

m_overall <- quap(
  alist(
    n_gaps ~ dpois(lambda),
    log(lambda) <- a + b * dph_c,
    a ~ dnorm(1, 0.5),
    b ~ dnorm(0, 0.05)
  ), data = d_overall, start = list(a = 0, b = 0)
)

d_sex <- list(n_gaps = E_fit_sex$n_gaps, sex_id = E_fit_sex$sex_id, dph_c = E_fit_sex$dph_c)

m_sex <- quap(
  alist(
    n_gaps ~ dpois(lambda),
    log(lambda) <- a[sex_id] + b[sex_id] * dph_c,
    a[sex_id] ~ dnorm(1, 0.5),
    b[sex_id] ~ dnorm(0, 0.1)
  ), data = d_sex, start = list(a = c(0,0), b = c(0,0))
)#
```


```{r posterior-frequency, include=FALSE}
post_f_freq <- link(m_sex, data = list(sex_id = rep(1, 100), dph_c = dph_seq_c_ov))
post_m_freq <- link(m_sex, data = list(sex_id = rep(2, 100), dph_c = dph_seq_c_ov))

mu_f <- apply(post_f_freq, 2, mean); PI_f <- apply(post_f_freq, 2, PI)
mu_m <- apply(post_m_freq, 2, mean); PI_m <- apply(post_m_freq, 2, PI)

plot_gaps_df <- data.frame(
  day   = DAY_SEQ,
  f_mu  = mu_f, f_low = PI_f[1,], f_high = PI_f[2,],
  m_mu  = mu_m, m_low = PI_m[1,], m_high = PI_m[2,]
)

post_s_samp <- extract.samples(m_sex)
gap_diff    <- exp(post_s_samp$a[,1]) - exp(post_s_samp$a[,2]) #
```


```{r results-frequency}
cat("Female slope:", round(mean(post_s_samp$b[,1]), 4),
    "PI:", round(PI(post_s_samp$b[,1]), 4),
    "P(<0):", mean(post_s_samp$b[,1] < 0), "\n")
cat("Male slope:  ", round(mean(post_s_samp$b[,2]), 4),
    "PI:", round(PI(post_s_samp$b[,2]), 4),
    "P(<0):", mean(post_s_samp$b[,2] < 0), "\n")
cat("Mean gap diff (F-M):", round(mean(gap_diff), 2),
    "PI:", round(PI(gap_diff), 2), "\n")
cat("P(Female > Male baseline):", mean(post_s_samp$a[,1] > post_s_samp$a[,2]), "\n")
```

Both sexes showed a consistent change in gap frequency across incubation. Because time is measured on a reversed scale (0 = hatching), the positive slopes indicate that gap frequency decreases as hatching approaches. Female gap frequency increased by 0.082 per day (89% PI [0.061, 0.104]; P(slope < 0) = 0), and males showed a very similar trend of 0.078 per day (89% PI [0.059, 0.098]; P(slope < 0) = 0). This indicates that neglect becomes less frequent closer to hatching, reflecting increasing parental attentiveness in both sexes.

Males initiated slightly more gaps than females throughout incubation, but this difference was small and uncertain, with a baseline difference of −0.07 gaps per day (89% PI [−0.16, 0.02]; P(Female > Male) = 0.12). The credible interval overlaps zero, suggesting weak evidence for a consistent sex difference after filtering short gaps. The slopes of both sexes were nearly identical, indicating that any sex difference remains stable across incubation rather than diverging or converging.

```{r plot-frequency, fig.width=7, fig.height=5, fig.cap="Expected number of neglect gaps per day by sex across incubation", warning=FALSE}
col_female <- "darkorange"; col_male <- "navyblue"

col_female <- "darkorange" 
col_male   <- "navyblue" 

# 3. Create the Plot

# Symulujemy obserwacje (n_gaps) dla każdego punktu w oryginalnych danych
# m_sex to Twój dopasowany model quap
post_samples <- sim(m_sex, data = d_sex)
# Bierzemy losową próbkę symulacji (np. pierwszą), aby pokazać realistyczny rozrzut
E_fit_sex$post_n_gaps <- post_samples[1, ]

# 2. Tworzenie wykresu
ggplot() +
  # --- 1. Wstęgi Niepewności ---
  geom_ribbon(data = plot_gaps_df, aes(x = day, ymin = f_low, ymax = f_high),  
              fill = col_female, alpha = 0.15) +
  geom_ribbon(data = plot_gaps_df, aes(x = day, ymin = m_low, ymax = m_high),  
              fill = col_male, alpha = 0.15) +
  
  # --- 2. Linie Trendu (Średnie) ---
  geom_line(data = plot_gaps_df, aes(x = day, y = f_mu), 
            color = col_female, linewidth = 1.2) +
  geom_line(data = plot_gaps_df, aes(x = day, y = m_mu), 
            color = col_male, linewidth = 1.2) +
  
  # --- 3. Bąbelki Posterior Predictive ---
  geom_count(data = E_fit_sex, 
             aes(x = day_prior_hatch, y = post_n_gaps, color = as.factor(sex_id)), 
             alpha = 0.3,
             position = position_jitter(width = 0.15, height = 0.05)) +
  
  # --- 4. Przywrócone Adnotacje Płci ---
  annotate("text", x = 27.8, y = max(plot_gaps_df$m_mu) * 1.4, 
           label = "\u2642", size = 10, color = col_male, fontface = "bold", family = "sans") +
  annotate("text", x = 27.8, y = 0.7, 
           label = "\u2640", size = 10, color = col_female, fontface = "bold", family = "sans") +
  
  # --- 5. Skalowanie i Etykiety ---
  scale_color_manual(values = c("1" = col_female, "2" = col_male), guide = "none") +
  scale_x_reverse(breaks = seq(0, 30, by = 5)) +
  scale_y_continuous(
    # Zoom na istotny zakres (model Poissona rzadko wychodzi poza 4-5)
    limits = c(-0.3, 2.2), 
    breaks = seq(0, 5, by = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_size_continuous(name = "n obs", range = c(1, 7)) +
  labs(title = "A)", x = "Days prior hatch", y = "Number of neglect gaps") +
  
  # --- 6. Motyw Graficzny ---
  theme_bw(base_size = 12, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    legend.position = c(0.90, 0.80),
    legend.background = element_blank(),
    axis.title = element_text(size = 13, color = "black"),
    axis.text = element_text(size = 11, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )
```

## Probability of any neglect

Not every observation day results in neglect. We therefore modelled the
probability that any neglect occurred on a given day using a binomial GLM
with a logit link, allowing intercepts and slopes to vary by sex.

$$\begin{aligned}
Y_i &\sim \text{Binomial}(1,\, p_i) \\
\text{logit}(p_i) &= \alpha_{sex[i]} + \beta_{sex[i]} \cdot \text{dph}_c \\
\alpha_j &\sim \text{Normal}(0,\, 1) \\
\beta_j  &\sim \text{Normal}(0,\, 0.2)
\end{aligned}$$

Where $Y_i$ is a binary indicator of whether any neglect was recorded on
observation $i$, and $p_i$ is the probability of neglect. The prior on
$\alpha$ is weakly informative on the logit scale, placing most prior
weight on probabilities between 0.1 and 0.9.


```{r prepare-occupancy-data, include=FALSE}
D_occ <- S_filt %>%
  filter(
    !is.na(sum_neglect_sec),
    !is.na(day_prior_hatch),
    !is.na(sx),
    day_prior_hatch >= 2,
    day_prior_hatch <= 28
  ) %>%
  mutate(
    sex_id = ifelse(sx == "f", 1, 2),
    dph_c = day_prior_hatch - mean(day_prior_hatch),
    any_neglect = ifelse(sum_neglect_sec > 0, 1, 0)
  )

d_occ <- list(
  y = D_occ$any_neglect,
  sex_id = D_occ$sex_id,
  dph_c = D_occ$dph_c
)
```

```{r fit-occupancy-model, include=FALSE, cache=TRUE}
m_occ <- quap(
  alist(
    y ~ dbinom(1, p),
    logit(p) <- a[sex_id] + b[sex_id] * dph_c,
    a[sex_id] ~ dnorm(0, 1),
    b[sex_id] ~ dnorm(0, 0.2)
  ),
  data = d_occ, start = list(a = c(0,0), b = c(0,0))
)#
```

```{r prior-check-occupancy, fig.width=6, fig.height=4, fig.cap="Prior predictive simulation for probability of neglect model"}
set.seed(43)
n_sim    <- 100
dph_grid <- seq(min(D_occ$dph_c), max(D_occ$dph_c), length.out = 100)

prior_f_occ <- matrix(NA, nrow = n_sim, ncol = length(dph_grid))
prior_m_occ <- matrix(NA, nrow = n_sim, ncol = length(dph_grid))

for(i in 1:n_sim){
  a <- rnorm(2, 0, 1)
  b <- rnorm(2, 0, 0.2)
  prior_f_occ[i,] <- inv_logit(a[1] + b[1] * dph_grid)
  prior_m_occ[i,] <- inv_logit(a[2] + b[2] * dph_grid)
}

par(mfrow = c(1, 2))
matplot(dph_grid, t(prior_f_occ), type = "l", lty = 1,
        col = col.alpha("darkorange", 0.3),
        ylim = c(0, 1),
        xlab = "dph (centered)", ylab = "P(neglect)",
        main = "Prior: Female")
matplot(dph_grid, t(prior_m_occ), type = "l", lty = 1,
        col = col.alpha("navyblue", 0.3),
        ylim = c(0, 1),
        xlab = "dph (centered)", ylab = "P(neglect)",
        main = "Prior: Male")
par(mfrow = c(1, 1))
```


```{r posterior-occupancy, include=FALSE}
post_f_occ <- link(m_occ, data = list(sex_id = rep(1, 100), dph_c = dph_seq_c_ov))
post_m_occ <- link(m_occ, data = list(sex_id = rep(2, 100), dph_c = dph_seq_c_ov))

mu_f_occ  <- apply(post_f_occ, 2, mean); PI_f_occ <- apply(post_f_occ, 2, PI)
mu_m_occ  <- apply(post_m_occ, 2, mean); PI_m_occ <- apply(post_m_occ, 2, PI)

plot_occ <- data.frame(
  day    = DAY_SEQ,
  f_mu   = mu_f_occ, f_low = PI_f_occ[1,], f_high = PI_f_occ[2,],
  m_mu   = mu_m_occ, m_low = PI_m_occ[1,], m_high = PI_m_occ[2,]
)

post_occ_samp <- extract.samples(m_occ)
prob_f    <- inv_logit(post_occ_samp$a[,1])
prob_m    <- inv_logit(post_occ_samp$a[,2])
prob_diff <- prob_f - prob_m
#
```

```{r results-occupancy}
cat("Female slope:", round(mean(post_occ_samp$b[,1]), 4),
    " PI:", round(PI(post_occ_samp$b[,1]), 4),
    " P(<0):", round(mean(post_occ_samp$b[,1] < 0), 3), "\n")
cat("Male slope:  ", round(mean(post_occ_samp$b[,2]), 4),
    " PI:", round(PI(post_occ_samp$b[,2]), 4),
    " P(<0):", round(mean(post_occ_samp$b[,2] < 0), 3), "\n")
cat("Mean prob diff (F-M):", round(mean(prob_diff), 3),
    " PI:", round(PI(prob_diff), 3), "\n")
cat("P(Female > Male baseline):", round(mean(post_occ_samp$a[,1] > post_occ_samp$a[,2]), 3), "\n")
#
```

The probability of neglect occurring on any given day showed the same pattern. With the reversed time scale, the positive slopes (female: 0.061; male: 0.067) indicate that the probability of neglect declines toward hatching (female PI [0.026, 0.096], P(slope < 0) = 0.002; male PI [0.036, 0.099], P(slope < 0) = 0). Thus, neglect was more common earlier in incubation and became less likely as hatching approached.

Males were credibly more likely to neglect than females, with a baseline difference of −0.074 (89% PI [−0.137, −0.012]; P(Female > Male) = 0.029). Although smaller than in the unfiltered data, this still supports a male-biased tendency to neglect at the daily level.

```{r plot-occupancy, fig.width=7, fig.height=5, fig.cap="Probability of any neglect per day by sex across incubation"}
plot_occ <- data.frame(
  day = DAY_SEQ,
  f_mu = mu_f_occ, f_low = PI_f_occ[1,], f_high = PI_f_occ[2,],
  m_mu = mu_m_occ, m_low = PI_m_occ[1,], m_high = PI_m_occ[2,]
)

post_samples_occ <- sim(m_occ, data = d_occ)#


D_occ$post_any_neglect <- post_samples_occ[1, ]

ggplot() +

  geom_ribbon(data = plot_occ,
              aes(x = day, ymin = f_low, ymax = f_high),
              fill = col_female, alpha = 0.15) +
  geom_ribbon(data = plot_occ,
              aes(x = day, ymin = m_low, ymax = m_high),
              fill = col_male, alpha = 0.15) +
  
  # Mean Lines
  geom_line(data = plot_occ,
            aes(x = day, y = f_mu),
            color = col_female, linewidth = 1.2) +
  geom_line(data = plot_occ,
            aes(x = day, y = m_mu),
            color = col_male, linewidth = 1.2) +
  
  # Bubble Points - TERAZ Z POSTERIOR PREDICTIVE (0 lub 1)
  geom_count(data = D_occ,
             aes(x = day_prior_hatch,
                 y = post_any_neglect, # Zmienione na dane symulowane
                 color = as.factor(sex_id)),
             alpha = 0.3,
             # Dodajemy jitter, żeby bąbelki na 0 i 1 nie były "płaskimi" plackami
             position = position_jitter(width = 0.15, height = 0.03)) +
  
  # Manual Color Scales
  scale_color_manual(values = c("1" = col_female, "2" = col_male),
                     guide = "none") +
  
  # Sex Labels
  annotate("text", x = 27.8, y = 0.7,
           label = "\u2642", size = 10,
           color = col_male, fontface = "bold") +
  annotate("text", x = 27.8, y = 0.20,
           label = "\u2640", size = 10,
           color = col_female, fontface = "bold") +
  
  # Scaling and Labels
  scale_x_reverse(breaks = seq(0, 30, by = 5)) +
  # Dodajemy mały margines w coord_cartesian, żeby jitter nie "uciekał" poza oś
  coord_cartesian(ylim = c(-0.1, 1.1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_size_continuous(name = "n obs", range = c(1, 6)) +
  labs(title = "C)", x = "Days prior hatch", y = "Probability of neglect") +
  
  # Theme
  theme_bw(base_size = 12, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    legend.position = c(0.90, 0.81),
    legend.background = element_blank(),
    legend.key = element_blank(),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11, color = "black"),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    plot.margin = margin(10, 10, 10, 10)
  )
```

---

## Duration of neglect (given neglect occurred)

To capture how long neglect episodes last when they do occur, we modelled
total neglect duration on the log scale using a Gaussian GLM, restricted
to days where neglect was recorded. This forms the second component of the
hurdle model — separating the question of *whether* neglect happens from
*how severe* it is when it does.

$$\begin{aligned}
\log(D_i) &\sim \text{Normal}(\mu_i,\, \sigma) \\
\mu_i &= \alpha_{sex[i]} + \beta_{sex[i]} \cdot \text{dph}_c \\
\alpha_j &\sim \text{Normal}(6,\, 0.5) \\
\beta_j  &\sim \text{Normal}(0,\, 0.1) \\
\sigma   &\sim \text{Exponential}(1)
\end{aligned}$$

Where $D_i$ is total neglect duration in seconds on observation $i$.
The prior on $\alpha$ is centered at 6 on the log scale, corresponding
to a baseline duration of roughly 400 seconds (~7 minutes), which is
biologically plausible for a seabird of this size.

```{r prepare-duration-data, include=FALSE}
D_dur <- S_filt %>%
  filter(
    !is.na(sum_neglect_sec),
    sum_neglect_sec > 0,
    !is.na(day_prior_hatch),
    !is.na(sx),
    day_prior_hatch >= 2,
    day_prior_hatch <= 28
  ) %>%
  mutate(
    sex_id = ifelse(sx == "f", 1, 2),
    dph_c = day_prior_hatch - mean(day_prior_hatch),
    neglect_log = log(sum_neglect_sec)
  )

d_dur <- list(
  y = D_dur$neglect_log,
  sex_id = D_dur$sex_id,
  dph_c = D_dur$dph_c
)
```


```{r fit-duration-model, include=FALSE, cache=TRUE}
m_dur <- quap(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- a[sex_id] + b[sex_id] * dph_c,
    a[sex_id] ~ dnorm(6, 0.2),
    b[sex_id] ~ dnorm(0, 0.05),
    sigma ~ dexp(1)
  ),
  data = d_dur, start = list(a = c(6,6), b = c(0,0), sigma = 1)
)
```

```{r prior-check-duration, fig.width=6, fig.height=4, fig.cap="Prior predictive simulation for neglect duration model"}
set.seed(42)
dph_grid_dur <- seq(min(D_dur$dph_c), max(D_dur$dph_c), length.out = 100)

prior_f_dur <- matrix(NA, nrow = n_sim, ncol = length(dph_grid_dur))
prior_m_dur <- matrix(NA, nrow = n_sim, ncol = length(dph_grid_dur))

for(i in 1:n_sim){
  a <- rnorm(2, 6, 0.2)
  b <- rnorm(2, 0, 0.05)
  prior_f_dur[i,] <- a[1] + b[1] * dph_grid_dur  # no exp() here
  prior_m_dur[i,] <- a[2] + b[2] * dph_grid_dur  # no exp() here
}

par(mfrow = c(1, 2))
matplot(dph_grid_dur, t(prior_f_dur), type = "l", lty = 1,
        col = col.alpha("darkorange", 0.3),
        ylim = c(4, 8),          # log scale limits (exp(12) ~ 160,000 sec)
        xlab = "dph (centered)", ylab = "log Duration (sec)",
        main = "Prior: Female")
matplot(dph_grid_dur, t(prior_m_dur), type = "l", lty = 1,
        col = col.alpha("navyblue", 0.3),
        ylim = c(4, 8),
        xlab = "dph (centered)", ylab = "log Duration (sec)",
        main = "Prior: Male")
par(mfrow = c(1, 1))
```


```{r posterior-duration, include=FALSE}
post_f_dur <- exp(link(m_dur, data = list(sex_id = rep(1, 100), dph_c = dph_seq_c_ov)))
post_m_dur <- exp(link(m_dur, data = list(sex_id = rep(2, 100), dph_c = dph_seq_c_ov)))

mu_f_dur  <- apply(post_f_dur, 2, mean); PI_f_dur <- apply(post_f_dur, 2, PI)
mu_m_dur  <- apply(post_m_dur, 2, mean); PI_m_dur <- apply(post_m_dur, 2, PI)

plot_dur_df <- data.frame(
  day    = DAY_SEQ,
  f_mu   = mu_f_dur, f_low = PI_f_dur[1,], f_high = PI_f_dur[2,],
  m_mu   = mu_m_dur, m_low = PI_m_dur[1,], m_high = PI_m_dur[2,]
)

post_dur_samp <- extract.samples(m_dur)
dur_f    <- exp(post_dur_samp$a[,1])
dur_m    <- exp(post_dur_samp$a[,2])
dur_diff <- dur_f - dur_m
```

```{r results-duration}
cat("Female slope:", round(mean(post_dur_samp$b[,1]), 4),
    " PI:", round(PI(post_dur_samp$b[,1]), 4),
    " P(<0):", round(mean(post_dur_samp$b[,1] < 0), 3), "\n")
cat("Male slope:  ", round(mean(post_dur_samp$b[,2]), 4),
    " PI:", round(PI(post_dur_samp$b[,2]), 4),
    " P(<0):", round(mean(post_dur_samp$b[,2] < 0), 3), "\n")
cat("Mean duration diff (F-M, sec):", round(mean(dur_diff), 1),
    " PI:", round(PI(dur_diff), 1), "\n")
cat("P(Female > Male baseline):", round(mean(post_dur_samp$a[,1] > post_dur_samp$a[,2]), 3), "\n")
```


When neglect did occur, its duration also changed over time. Again, given the reversed scale, the positive slopes (female: 0.0633; male: 0.0638) indicate that neglect duration decreases toward hatching (female PI [0.0344, 0.0934]; male PI [0.0368, 0.0915]; both P(slope < 0) = 0). This shows that not only does neglect become less frequent, but individual events also become shorter as hatching approaches.

There was no credible sex difference in duration. The mean baseline difference was −38 seconds (89% PI [−140.1, 62.7]; P(Female > Male) = 0.275), indicating substantial overlap between sexes.

Taken together, the hurdle model reveals an important asymmetry: the sex difference in neglect is primarily driven by whether they neglect at all, rather than by differences in how long neglect lasts once it begins. Both components decline toward hatching, pointing to a shared increase in parental commitment as the critical hatching window approaches.


```{r plot-duration, fig.width=7, fig.height=5, fig.cap="Total neglect duration (seconds, log scale) given neglect occurred, by sex across incubation"}
post_samples_log_dur <- sim(m_dur, data = d_dur)


D_dur$post_sum_neglect_sec <- exp(post_samples_log_dur[1, ])

plot_df <- data.frame(
  day = DAY_SEQ,
  f_mu = mu_f_dur, f_low = PI_f_dur[1,], f_high = PI_f_dur[2,],
  m_mu = mu_m_dur, m_low = PI_m_dur[1,], m_high = PI_m_dur[2,]
)

ggplot() +
  geom_ribbon(data = plot_df,
              aes(x = day, ymin = f_low, ymax = f_high),
              fill = col_female, alpha = 0.15) +
  geom_ribbon(data = plot_df,
              aes(x = day, ymin = m_low, ymax = m_high),
              fill = col_male, alpha = 0.15) +
  
  # 2. Regression Lines
  geom_line(data = plot_df,
            aes(x = day, y = f_mu),
            color = col_female, linewidth = 1.2) +
  geom_line(data = plot_df,
            aes(x = day, y = m_mu),
            color = col_male, linewidth = 1.2) +
  
  # 3. Bubble Points - TERAZ Z POSTERIOR PREDICTIVE
  # geom_count automatycznie pogrupuje podobne wartości symulowane
  geom_count(data = D_dur,
             aes(x = day_prior_hatch,
                 y = post_sum_neglect_sec, 
                 color = as.factor(sex_id)),
             alpha = 0.3) + 
  
  # 4. Sex Annotations
  annotate("text", x = 27.8, y = plot_df$m_mu[100] * 2.4, 
           label = "\u2642", size = 10, color = col_male, fontface = "bold", family = "sans") +
  annotate("text", x = 27.8, y = plot_df$f_mu[100] * 0.4, 
           label = "\u2640", size = 10, color = col_female, fontface = "bold", family = "sans") +

# 5. Scales and Labels
scale_color_manual(values = c("1" = col_female, "2" = col_male), guide = "none") +
  
  scale_size_continuous(range = c(0.5, 2), guide = "none") + 
  
  scale_y_log10(
    breaks = c(100, 1000, 10000, 100000), 
    labels = scales::label_comma(),
    expand = expansion(mult = c(0.1, 0.1))
  ) +
  scale_x_reverse(breaks = seq(0, 30, by = 5)) +
  labs(title = "B)", 
       x = "Days prior hatch",
       y = "Duration of neglect events (sec, log scale)") +
  
  # 7. Theme
  theme_bw(base_size = 12, base_family = "sans") +
  theme(
    panel.grid = element_blank(),
    legend.position = "none", # Najpewniejszy sposób na usunięcie wszystkich legend
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 11, color = "black"),
    plot.margin = margin(10, 10, 10, 10)
  )

```


# Neglect behaviour in relation to pair status and similarity

In this section we examine whether pair experience (new vs. established
pairs) and pair similarity predict neglect behaviour. We model both gap
frequency and total neglect duration for each predictor separately, using
a hurdle approach — the frequency model includes all observations, while
the duration model is restricted to days where neglect was recorded.

---

## Pair status

We modelled gap frequency using a Poisson GLM and neglect duration using
a Normal GLM on the log scale, both with pair status (New vs. Old) as a
binary predictor and session as an index variable to account for repeated
measures across incubation. Due to low sample size, all three sessions
were included in this analysis.

$$\begin{aligned}
N_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \alpha_{pair[i]} + \beta_1 \cdot dph\_c_i \\
\alpha_j &\sim \text{Normal}(0,\, 0.8) \\
\beta_1 &\sim \text{Normal}(0,\, 0.5)
\end{aligned}$$

$$\begin{aligned}
\log(D_i) &\sim \text{Normal}(\mu_i,\, \sigma) \\
\mu_i &= \alpha_{pair[i]} + \beta_1 \cdot dph\_c_i \\
\alpha_j &\sim \text{Normal}(6,\, 0.5) \\
\beta_1 &\sim \text{Normal}(0,\, 0.5) \\
\sigma &\sim \text{Exponential}(1)
\end{aligned}$$
Where $\text{pair}_{01}$ is a binary indicator (0 = New, 1 = Old) and
session $j$ captures variation across incubation stages.

```{r prepare-pairstatus-data, include=FALSE}
E <- readRDS("./EDA_egg_neglect_index_neglect_table.RDS")
Y <- readRDS("./EDA_egg_neglect_index_gaps_table.RDS")

# --- Filter short gaps and rebuild E summaries ---
Y_filtered <- bind_rows(Y) %>%          # flatten list to single dataframe
  filter(dur >= 300)                     # keep only gaps >= 5 min

Y_summary <- Y_filtered %>%
  group_by(season, session, nest) %>%
  summarise(
    n_gaps    = n(),
    sum_gaps  = sum(dur),
    mean_gaps = mean(dur),
    .groups   = "drop"
  )

E <- E %>%
  select(-n_gaps, -sum_gaps, -mean_gaps) %>%
  left_join(Y_summary, by = c("season", "session", "nest")) %>%
  mutate(
    n_gaps    = replace_na(n_gaps,    0),
    sum_gaps  = replace_na(sum_gaps,  0),
    mean_gaps = replace_na(mean_gaps, 0)
  )

nest_summary <- E %>%
  group_by(season, Pair_status) %>%
  summarise(n_distinct_nests = n_distinct(nest), .groups = 'drop')

E <- E %>%
  mutate(
    pair_01     = ifelse(Pair_status == "Old", 2, 1),
    has_neglect = ifelse(sum_gaps > 0, 1, 0)
  )

E <- E %>%
  mutate(
    session = case_when(
      day_prior_hatch >= 21 & day_prior_hatch <= 30 ~ "early",
      day_prior_hatch >= 11 & day_prior_hatch <= 20 ~ "mid",
      day_prior_hatch >= 0  & day_prior_hatch <= 10 ~ "late",
      TRUE ~ NA_character_),
    dph_c = day_prior_hatch - mean(day_prior_hatch)
  )

# Filter global datasets
D_all     <- E %>% filter(!is.na(Pair_status), !is.na(n_gaps), !is.na(session))
D_nonzero <- D_all %>% filter(sum_gaps > 0)
```

```{r prior-check-pairstatus, fig.width=6, fig.height=4, fig.cap="Prior predictive simulation for pair status models"}
set.seed(42)
n_sim <- 100

prior_freq_new <- prior_freq_old <- numeric(n_sim)
prior_dur_new  <- prior_dur_old  <- numeric(n_sim)

for(i in 1:n_sim){
  a   <- rnorm(1, 0, 0.8)
  b   <- rnorm(1, 0, 0.5)
  prior_freq_new[i] <- exp(a)
  prior_freq_old[i] <- exp(a + b)

  a_d <- rnorm(1, 6, 0.5)
  b_d <- rnorm(1, 0, 0.5)
  prior_dur_new[i] <- a_d
  prior_dur_old[i] <- a_d + b_d
}

par(mfrow = c(1, 2))
boxplot(list(New = prior_freq_new, Old = prior_freq_old),
        col = c("gray80", "gray60"),
        ylab = "Expected n gaps", main = "Prior: Frequency",
        outline = FALSE)
boxplot(list(New = prior_dur_new, Old = prior_dur_old),
        col = c("gray80", "gray60"),
        ylab = "log Duration (sec)", main = "Prior: Duration (log scale)",
        outline = FALSE)
par(mfrow = c(1, 1))
```

```{r fit-pairstatus-models, include=FALSE, cache=TRUE}
m_status_freq <- quap(
  alist(
    n_gaps ~ dpois(lambda),
    log(lambda) <- a[pair_01] + b* dph_c,
    a[pair_01] ~ dnorm(0, 0.8),
    b ~ dnorm(0, 0.5)
  ), data = D_all
)

m_status_dur <- quap(
  alist(
    sum_gaps ~ dnorm(mu, sigma),
    mu <- a[pair_01] + b* dph_c,
    a[pair_01] ~ dnorm(6, 0.5),
    b ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = D_nonzero
)

post_f <- extract.samples(m_status_freq)
post_d <- extract.samples(m_status_dur)
```

```{r posterior-pairstatus, include=FALSE}
plot_freq_df <- data.frame(
  status = factor(c("New", "Old"), levels = c("New", "Old")),
  mu   = c(mean(exp(post_f$a)), mean(exp(post_f$a + post_f$b))),
  low  = c(PI(exp(post_f$a))[1], PI(exp(post_f$a + post_f$b))[1]),
  high = c(PI(exp(post_f$a))[2], PI(exp(post_f$a + post_f$b))[2])
)

plot_dur_df <- data.frame(
  status = factor(c("New", "Old"), levels = c("New", "Old")),
  mu   = c(mean(exp(post_d$a)), mean(exp(post_d$a + post_d$b))),
  low  = c(PI(exp(post_d$a))[1], PI(exp(post_d$a + post_d$b))[1]),
  high = c(PI(exp(post_d$a))[2], PI(exp(post_d$a + post_d$b))[2])
)


# plot posteriors
f_sim <- sim(m_status_freq, data = list(pair_01 = 1:2, dph_c = c(0,0)), n = 2000)
d_sim <- sim(m_status_dur,  data = list(pair_01 = 1:2, dph_c = c(0,0)), n = 2000)

df_post_f <- data.frame(
  Pair_status = rep(c("New", "Old"), each = 2000),
  val = as.vector(f_sim)
)

df_post_d <- data.frame(
  Pair_status = rep(c("New", "Old"), each = 2000),
  val = as.vector(d_sim)
)
```

```{r results-pairstatus}
post_f <- extract.samples(m_status_freq)
post_d <- extract.samples(m_status_dur)

cat("\n--- PAIR STATUS RESULTS (Global Intercept) ---\n")
cat("FREQ: Rate ratio (Old/New):", round(mean(exp(post_f$b)), 3), " | 89% PI:", round(PI(exp(post_f$b)), 3), "\n")
cat("DUR:  Duration Ratio (Old/New):", round(mean(exp(post_d$b)), 3), " | 89% PI:", round(PI(exp(post_d$b)), 3), "\n")
cat("DUR:  P(Old Duration > New Duration):", round(mean(post_d$b > 0), 3), "\n")

```

Pair experience significantly influenced the frequency of nest neglect, though its effect on the duration of neglect was less clear. Established pairs (Old) initiated neglect events more frequently than new pairs, with a rate ratio of 1.21 (89% PI [1.16, 1.25]). These results provide strong evidence that experienced pairs leave the nest more often than new pairs throughout the incubation period.

However, once neglect occurred, there was no credible evidence that established pairs stayed off the nest for longer durations than new pairs. Although the mean duration ratio was 1.15 in favor of Old pairs, the credible interval was very wide (89% PI [0.45, 2.23]) and the posterior probability that Old pairs had longer durations was only 0.514. This indicates that while experienced pairs neglect the nest more often, the length of those individual neglect episodes does not differ significantly based on pair status.

```{r plot-pairstatus, fig.width=8, fig.height=5, fig.cap="Gap frequency (A) and neglect duration (B) by pair status. Points show model estimates with 89% PI; violins and jittered points show raw data."}
df_post_d_clean <- df_post_d
df_post_d_clean$val[df_post_d_clean$val <= 0] <- 1 # Zamiana na 1s, by log10 działał

# --- Wykres A: Bez zmian (Poisson nie ma problemu z log) ---
p_freq <- ggplot() +
  geom_jitter(data = D_all, aes(x = factor(pair_01, labels=c("New", "Old")), y = n_gaps),
              width = 0.2, alpha = 0.2, size = 1, color = "gray70") +
  geom_violin(data = df_post_f, aes(x = Pair_status, y = val), 
              fill = "gray95", color = "black", linewidth = 0.5, alpha = 0.5) +
  geom_errorbar(data = plot_freq_df, aes(x = status, ymin = low, ymax = high), 
                width = 0.05, linewidth = 1, color = "black") +
  geom_point(data = plot_freq_df, aes(x = status, y = mu), 
             size = 3.5, color = "black") +
  coord_cartesian(ylim = c(0, 5)) +
  labs(title = "A) Gap Frequency", x = "Pair Status", y = "Number of gaps") +
  theme_classic(base_size = 12)

# --- Wykres B: Poprawiony pod skalę logarytmiczną ---
p_dur <- ggplot() +
  geom_jitter(data = D_nonzero, aes(x = factor(pair_01, labels=c("New", "Old")), y = sum_gaps),
              width = 0.2, alpha = 0.2, size = 1, color = "gray70") +
  # Używamy wyczyszczonych danych (df_post_d_clean)
  geom_violin(data = df_post_d_clean, aes(x = Pair_status, y = val), 
              fill = "gray95", color = "black", linewidth = 0.5, alpha = 0.5) +
  geom_errorbar(data = plot_dur_df, aes(x = status, ymin = low, ymax = high), 
                width = 0.05, linewidth = 1, color = "black") +
  geom_point(data = plot_dur_df, aes(x = status, y = mu), 
             size = 3.5, color = "black") +
  # Ustawienie limitów wewnątrz skali, by uniknąć wyrzucania punktów przez jitter
  scale_y_log10(labels = scales::label_comma(), 
                breaks = c(1, 100, 1000, 10000, 100000)) +
  labs(title = "B) Neglect Duration", x = "Pair Status", y = "Duration (sec)") +
  theme_classic(base_size = 12)


# Połączenie wykresów
library(patchwork)
p_freq + p_dur

```

Differences in nest neglect frequency and duration between new and established pairs. A) Number of gaps initiated per observation session. B) Total duration of neglect episodes (seconds, plotted on log scale) for sessions where neglect occurred. In both panels, large red points represent posterior means, and black error bars represent 89% posterior credible intervals derived from Bayesian GLMs (Poisson for frequency, log-Normal for duration). The shaded violin plots illustrate the full posterior distribution of the expected values. Small gray jittered points show the raw observed data for each pair status. While established pairs (Old) initiated neglect gaps significantly more frequently than new pairs, the duration of individual neglect episodes did not differ credibly between pair types.

---

## Pair similarity

We modelled gap frequency and neglect duration as a function of a
continuous pair Similarity Index, using only the first incubation session
as it showed the highest number of exits. No session index was needed
here as only one session was included.

$$\begin{aligned}
N_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \alpha + \beta \cdot \text{Similarity}_i \\
\alpha &\sim \text{Normal}(0,\, 0.8) \\
\beta  &\sim \text{Normal}(0,\, 0.5)
\end{aligned}$$

$$\begin{aligned}
\log(D_i) &\sim \text{Normal}(\mu_i,\, \sigma) \\
\mu_i &= \alpha + \beta \cdot \text{Similarity}_i \\
\alpha &\sim \text{Normal}(6,\, 0.5) \\
\beta  &\sim \text{Normal}(0,\, 0.5) \\
\sigma &\sim \text{Exponential}(1)
\end{aligned}$$

```{r prepare-similarity-data, include=FALSE}
D_sim         <- D_all %>% filter(!is.na(Similarity_Index), session == "early")
D_sim_nonzero <- D_sim %>% filter(sum_gaps > 0) %>% mutate(y = log(sum_gaps))
```

```{r prior-check-similarity, fig.width=6, fig.height=4, fig.cap="Prior predictive simulation for pair similarity models"}
set.seed(42)
xseq_prior <- seq(0, 1, length.out = 100)

prior_sim_freq <- matrix(NA, nrow = n_sim, ncol = 100)
prior_sim_dur  <- matrix(NA, nrow = n_sim, ncol = 100)

for(i in 1:n_sim){
  a   <- rnorm(1, 0, 0.8); b   <- rnorm(1, 0, 0.5)
  prior_sim_freq[i,] <- exp(a + b * xseq_prior)

  a_d <- rnorm(1, 6, 0.5); b_d <- rnorm(1, 0, 0.5)
  prior_sim_dur[i,]  <- a_d + b_d * xseq_prior  # log scale
}

par(mfrow = c(1, 2))
matplot(xseq_prior, t(prior_sim_freq), type = "l", lty = 1,
        col = col.alpha("black", 0.3),
        xlab = "Similarity Index", ylab = "Expected n gaps",
        main = "Prior: Frequency")
matplot(xseq_prior, t(prior_sim_dur), type = "l", lty = 1,
        col = col.alpha("black", 0.3),
        xlab = "Similarity Index", ylab = "log Duration (sec)",
        main = "Prior: Duration (log scale)")
par(mfrow = c(1, 1))
```

```{r fit-similarity-models, include=FALSE, cache=TRUE}
m_sim_freq <- quap(
  alist(
    n_gaps ~ dpois(lambda),
    log(lambda) <- a + b * Similarity_Index,
    a ~ dnorm(0, 0.8),
    b ~ dnorm(0, 0.5)
  ), data = D_sim
)

m_sim_dur <- quap(
  alist(
    y ~ dnorm(mu, sigma),
    mu <- a + b * Similarity_Index,
    a ~ dnorm(6, 0.5),
    b ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data = D_sim_nonzero
)
```

```{r posterior-similarity, include=FALSE, cache = TRUE}
post_sim_f <- extract.samples(m_sim_freq)
post_sim_d <- extract.samples(m_sim_dur)

xseq <- seq(0, 1, length.out = 100)

link_f <- link(m_sim_freq, data = list(Similarity_Index = xseq))
link_d <- exp(link(m_sim_dur, data = list(Similarity_Index = xseq)))

plot_sim_f <- data.frame(
  x    = xseq,
  mu   = apply(link_f, 2, mean),
  low  = apply(link_f, 2, PI, prob = 0.89)[1,],
  high = apply(link_f, 2, PI, prob = 0.89)[2,]
)

plot_sim_d <- data.frame(
  x    = xseq,
  mu   = apply(link_d, 2, mean),
  low  = apply(link_d, 2, PI, prob = 0.89)[1,],
  high = apply(link_d, 2, PI, prob = 0.89)[2,]
)
```

```{r results-similarity}
cat("FREQ: Slope b:", round(mean(post_sim_f$b), 3),
    " 89% PI:", round(PI(post_sim_f$b), 3), "\n")
cat("FREQ: P(slope > 0):", round(mean(post_sim_f$b > 0), 3), "\n")
cat("DUR:  Slope b:", round(mean(post_sim_d$b), 3),
    " 89% PI:", round(PI(post_sim_d$b), 3), "\n")
cat("DUR:  P(slope > 0):", round(mean(post_sim_d$b > 0), 3), "\n")
```

Gap frequency increased credibly with pair similarity, with a slope of
0.542 on the log scale (89% PI [0.341, 0.740]; P(slope > 0) = 1.0),
indicating that more similar pairs initiated more frequent nest exits.
This is a strong and consistent effect with no posterior probability
supporting a negative relationship.

Neglect duration showed a weaker and less credible positive trend with
similarity (slope: 0.360, 89% PI [−0.295, 1.023]; P(slope > 0) = 0.806),
with the interval spanning zero. While the direction is consistent with
the frequency result, the evidence is insufficient to conclude that more
similar pairs also show longer neglect episodes when they do occur.

Taken together, pair similarity appears to primarily influence how often
pairs neglect rather than how long those episodes last — a pattern that
mirrors the sex differences observed earlier, where frequency was the
more consistent axis of variation.

```{r plot-similarity, fig.width=8, fig.height=5, fig.cap="Gap frequency (A) and neglect duration (B) in relation to pair Similarity Index. Lines show posterior mean with 89% PI; bubble size indicates number of overlapping observations."}
shared_size_scale <- scale_size_continuous(
  name   = "n obs",
  range  = c(1.5, 8),
  limits = c(1, 10),
  breaks = c(1, 2, 5, 10)
)

p_sim_freq <- ggplot() +
  geom_count(data = D_sim, aes(x = Similarity_Index, y = n_gaps),
             alpha = 0.2, color = "gray40") +
  geom_ribbon(data = plot_sim_f, aes(x = x, ymin = low, ymax = high),
              fill = "black", alpha = 0.1) +
  geom_line(data = plot_sim_f, aes(x = x, y = mu),
            color = "black", linewidth = 1.2) +
  shared_size_scale +
  labs(title = "A)", x = "Similarity Index", y = "Number of exits") +
  theme_bw(base_size = 12, base_family = "sans") +
  theme(panel.grid = element_blank(),
        legend.background = element_blank())

p_sim_dur <- ggplot() +
  geom_count(data = D_sim_nonzero, aes(x = Similarity_Index, y = sum_gaps),
             alpha = 0.2, color = "gray40") +
  geom_ribbon(data = plot_sim_d, aes(x = x, ymin = low, ymax = high),
              fill = "black", alpha = 0.1) +
  geom_line(data = plot_sim_d, aes(x = x, y = mu),
            color = "black", linewidth = 1.2) +
  scale_y_log10(labels = scales::label_number()) +
  shared_size_scale +
  labs(title = "B)", x = "Similarity Index", y = "Duration of neglect (sec, log scale)") +
  theme_bw(base_size = 12, base_family = "sans") +
  theme(panel.grid = element_blank(),
        legend.background = element_blank())

p_sim_freq + p_sim_dur +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")
```

# Repeatability of neglect behaviour

To determine whether individuals are consistent in their tendency to
neglect across incubation sessions, we estimated the repeatability of
neglect occurrence using the `rptR` package. We modelled whether an
individual neglected at all on a given observation (binary: 0/1) using
a binomial repeatability model, with session as a fixed effect to
account for the systematic change in neglect across incubation stages,
and individual identity (`ringno`) as the grouping factor. This approach
asks: after accounting for when in incubation the observation was made,
is an individual consistently neglectful or not across all sessions in
which it was observed? Repeatability was estimated using 1000 bootstrap
iterations for confidence intervals and 1000 permutations for
significance testing.

$$\begin{aligned}
\text{is_neglecting}_i &\sim \text{Binomial}(1,\, p_i) \\
\text{logit}(p_i) &= \beta \cdot \text{dph}_c,i + u_{ringno[i]} \\
u_{ringno} &\sim \text{Normal}(0,\, \sigma^2_{ringno})
\end{aligned}$$

Where repeatability $R$ is defined as the proportion of total phenotypic
variance attributable to among-individual differences:

$$R = \frac{\sigma^2_{ringno}}{\sigma^2_{ringno} + \sigma^2_{residual}}$$

We estimated repeatability separately for the full dataset and for each
sex, to test whether the consistency of neglect tendency differs between
females and males.

```{r prepare-repeatability-data, include=FALSE, cache = TRUE}
library(rptR)

S <- readRDS("EDA_sex_neglect_wide.RDS")
S <- S %>%
  filter(!is.na(day_prior_hatch),
         day_prior_hatch >= 2,
         day_prior_hatch <= 28) %>%
  mutate(
    dph_c = day_prior_hatch - mean(day_prior_hatch, na.rm = TRUE),
    is_neglecting = ifelse(sum_neglect_sec > 0, 1, 0)
  )

S_male   <- S %>% filter(sx == "m")
S_female <- S %>% filter(sx == "f")
```

```{r fit-repeatability-overall, include=FALSE, cache=TRUE, message = FALSE}
rep_binary <- rpt(
  is_neglecting ~ dph_c + (1 | ringno),
  grname   = "ringno",
  data     = S,
  datatype = "Binary",
  nboot    = 1000,
  npermut  = 1000
)
```

```{r fit-repeatability-sex, include=FALSE, cache=TRUE, message = FALSE}
rep_binary_male <- rpt(
  is_neglecting ~ dph_c + (1 | ringno),
  grname   = "ringno",
  data     = S_male,
  datatype = "Binary",
  nboot    = 1000,
  npermut  = 1000
)

rep_binary_female <- rpt(
  is_neglecting ~ dph_c + (1 | ringno),
  grname   = "ringno",
  data     = S_female,
  datatype = "Binary",
  nboot    = 1000,
  npermut  = 1000
)
```

```{r results-repeatability, echo=FALSE}
library(knitr)

extract_rpt <- function(mod, name) {
  data.frame(
    Group   = name,
    R_link  = round(mod$R["R_link", "ringno"], 3),
    CI_link = paste0("[", round(mod$CI_emp$CI_link["ringno", "2.5%"],  3),
                     ", ", round(mod$CI_emp$CI_link["ringno", "97.5%"], 3), "]"),
    R_org   = round(mod$R["R_org", "ringno"], 3),
    CI_org  = paste0("[", round(mod$CI_emp$CI_org["ringno", "2.5%"],  3),
                     ", ", round(mod$CI_emp$CI_org["ringno", "97.5%"], 3), "]"),
    P_permut = mod$P["ringno", "P_permut_link"],
    stringsAsFactors = FALSE
  )
}

summary_table <- rbind(
  extract_rpt(rep_binary,        "Overall"),
  extract_rpt(rep_binary_male,   "Males"),
  extract_rpt(rep_binary_female, "Females")
)

kable(summary_table,
      col.names = c("Group", "R (Link)", "95% CI (Link)",
                    "R (Original)", "95% CI (Original)", "P (Permutation)"),
      caption   = "Repeatability of individual neglect tendency across
                   incubation sessions. Session included as fixed effect.",
      align     = c("l", "c", "c", "c", "c", "c"))
```


Neglect tendency showed significant and consistent repeatability across incubation sessions. In the overall sample, individuals were moderately repeatable in whether they neglected at all, with a link-scale repeatability of R = 0.247 (95% CI [0.116, 0.369]; P = 0.001 by permutation test), and an original-scale estimate of R = 0.250 (95% CI [0.118, 0.408]). This indicates that a meaningful proportion of variation in neglect occurrence is attributable to stable among-individual differences, rather than being purely driven by session or chance.

Repeatability was significant in both sexes, though point estimates suggest that females may show higher individual consistency than males. Females had a link-scale repeatability of R = 0.262 (95% CI [0.063, 0.408]), compared to R = 0.141 (95% CI [0.005, 0.301]) in males, with both significant at P < 0.005 by permutation test. The confidence intervals of both sexes overlap substantially, so caution is warranted in interpreting this difference as biologically meaningful.

Taken together, these results suggest that neglect tendency is a repeatable individual trait in this population — some individuals consistently neglect across incubation sessions while others consistently do not. 

# Effect of neglect on incubation duration

To test whether the cumulative amount of neglect during incubation
affects how long incubation lasts, we modelled incubation duration
(days) as a function of total neglect accumulated across all three
sessions. Only nests with complete data across all three incubation
sessions were included, to ensure total neglect was not underestimated
due to missing observations. Total neglect duration was log-transformed
prior to fitting to address its right-skewed distribution, and then
mean-centered.

$$\begin{aligned}
\text{Duration}_i &\sim \text{Normal}(\mu_i,\, \sigma) \\
\mu_i &= \alpha + \beta \cdot \log(\text{neglect})_{c,i} \\
\alpha &\sim \text{Normal}(29,\, 0.5) \\
\beta  &\sim \text{Normal}(0,\, 0.5) \\
\sigma &\sim \text{Exponential}(1)
\end{aligned}$$

Where $\log(\text{neglect})_{c,i}$ is the log-transformed total neglect
in seconds, mean-centered. The prior on $\alpha$ is centered at 29 days,
consistent with the known incubation period of Little Auk. The prior on
$\beta$ is centered at zero, reflecting no strong prior expectation about
the direction of the effect.

```{r prepare-incubation-data, include=FALSE}
E <- readRDS("./EDA_egg_neglect_index_neglect_table.RDS")

nest_gaps <- E %>%
  group_by(season, nest) %>%
  filter(n() == 3) %>%
  summarize(
    total_sum_gaps      = sum(sum_gaps, na.rm = TRUE),
    incubation_duration = first(Incubation_Duration),
    .groups = "drop"
  ) %>%
  mutate(
    log_neglect   = log(total_sum_gaps + 1),
    log_neglect_c = log_neglect - mean(log_neglect)
  ) %>%
  filter(!is.na(incubation_duration))

d_quap <- list(
  incubation_duration = nest_gaps$incubation_duration,
  log_neglect_c       = nest_gaps$log_neglect_c
)

log_neglect_seq <- seq(min(nest_gaps$log_neglect_c),
                       max(nest_gaps$log_neglect_c),
                       length.out = 50)
```

```{r prior-check-incubation, fig.width=6, fig.height=4, fig.cap="Prior predictive simulation for incubation duration model. Red dashed lines indicate the biologically plausible range of incubation duration in Little Auk."}
set.seed(123)
n_sim <- 50

plot(NULL, xlim = range(log_neglect_seq), ylim = c(26, 34),
     xlab = "log Total neglect (centered)",
     ylab = "Incubation duration (days)",
     main = "Prior predictive: incubation duration")
for(i in 1:n_sim){
  a <- rnorm(1, 29, 0.5)
  b <- rnorm(1, 0, 0.5)
  lines(log_neglect_seq, a + b * log_neglect_seq,
        col = col.alpha("grey", 0.6))
}
abline(h = c(28, 32), col = "red", lty = 2)
```

```{r fit-incubation-model, include=FALSE, cache=TRUE}
m_quap <- quap(
  alist(
    incubation_duration ~ dnorm(mu, sigma),
    mu    <- a + b * log_neglect_c,
    a     ~ dnorm(29, 0.5),
    b     ~ dnorm(0, 0.5),
    sigma ~ dexp(1)
  ),
  data  = d_quap,
  start = list(a = 29, b = 0, sigma = 1)
)
```

```{r posterior-incubation, include=FALSE}
post_mu  <- link(m_quap, data = list(log_neglect_c = log_neglect_seq))
mu_mean  <- apply(post_mu, 2, mean)
mu_PI    <- apply(post_mu, 2, PI, prob = 0.89)

post_sim <- sim(m_quap, data = list(log_neglect_c = log_neglect_seq))
pred_PI  <- apply(post_sim, 2, PI, prob = 0.89)
```

```{r results-incubation}
print(precis(m_quap))
post_inc <- extract.samples(m_quap)
cat("P(b > 0):", round(mean(post_inc$b > 0), 3), "\n")
cat("P(b < 0):", round(mean(post_inc$b < 0), 3), "\n")
```

Total neglect accumulated across incubation showed a weak positive
association with incubation duration, though the evidence was not
conclusive. The posterior slope was 0.155 days per unit increase in log
neglect (89% PI [−0.012, 0.322]; P(β > 0) = 0.927), with the interval
just crossing zero. The intercept confirmed a mean incubation duration
of 29.1 days (89% PI [28.9, 29.4]), consistent with the known biology
of the species.

While most posterior weight supports a positive relationship — suggesting
that nests accumulating more neglect tend toward slightly longer
incubation — the uncertainty is substantial and the effect size is small.
A one-unit increase in log neglect corresponds to a posterior mean
increase of only 0.16 days. These results provide weak and inconclusive
evidence that neglect prolongs incubation, and should be interpreted
cautiously given the marginal credibility of the effect.

```{r plot-incubation, fig.width=7, fig.height=5, fig.cap="Incubation duration as a function of log total neglect accumulated across all three sessions. Line shows posterior mean; dark shading shows 89% PI of the mean; light shading shows 89% prediction interval."}
ggplot() +
  geom_point(data = nest_gaps,
             aes(x = log_neglect_c, y = incubation_duration),
             color = "steelblue", alpha = 0.6, size = 2) +
  geom_ribbon(aes(x = log_neglect_seq,
                  ymin = pred_PI[1,], ymax = pred_PI[2,]),
              fill = "black", alpha = 0.05) +
  geom_ribbon(aes(x = log_neglect_seq,
                  ymin = mu_PI[1,], ymax = mu_PI[2,]),
              fill = "black", alpha = 0.15) +
  geom_line(aes(x = log_neglect_seq, y = mu_mean),
            linewidth = 1.2, color = "black") +
  labs(x = "log Total neglect, centered (seconds)",
       y = "Incubation duration (days)") +
  theme_bw(base_size = 12, base_family = "sans") +
  theme(panel.grid = element_blank())
```

# Seasonal differences in neglect behaviour

To examine whether neglect behaviour varied across breeding seasons, we
modelled gap frequency, probability of neglect, and neglect duration
separately for each season. All three models included season-specific
intercepts to allow the baseline level of neglect to vary between
seasons, while a shared slope captured the common trajectory of neglect
across incubation days.

## Gap frequency

$$\begin{aligned}
N_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \alpha_{season[i]} + \beta \cdot \text{dph}_c \\
\alpha_j &\sim \text{Normal}(1,\, 0.5) \\
\beta    &\sim \text{Normal}(0,\, 0.05)
\end{aligned}$$

## Probability of neglect

$$\begin{aligned}
Y_i &\sim \text{Binomial}(1,\, p_i) \\
\text{logit}(p_i) &= \alpha_{season[i]} + \beta \cdot \text{dph}_c \\
\alpha_j &\sim \text{Normal}(0,\, 1) \\
\beta    &\sim \text{Normal}(0,\, 0.2)
\end{aligned}$$

## Duration of neglect (given neglect occurred)

$$\begin{aligned}
\log(D_i) &\sim \text{Normal}(\mu_i,\, \sigma) \\
\mu_i &= \alpha_{season[i]} + \beta \cdot \text{dph}_c \\
\alpha_j &\sim \text{Normal}(6,\, 0.5) \\
\beta    &\sim \text{Normal}(0,\, 0.1) \\
\sigma   &\sim \text{Exponential}(1)
\end{aligned}$$

Where $\text{dph}_c$ is days prior to hatch, mean-centered. Season-specific
intercepts $\alpha_j$ capture differences in the overall level of neglect
between years, while the shared slope $\beta$ assumes a common trajectory
across incubation for all seasons.

```{r prepare-seasonal-data, include=FALSE}
E <- readRDS("./EDA_egg_neglect_index_neglect_table.RDS")

cols_season <- c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00")

E_fit_season <- E %>%
  filter(!is.na(day_prior_hatch), !is.na(n_gaps), !is.na(season),
         day_prior_hatch >= 2, day_prior_hatch <= 28) %>%
  mutate(
    dph_c     = day_prior_hatch - mean(day_prior_hatch),
    has_gap   = ifelse(sum_gaps > 0, 1, 0),
    season    = factor(season)
  )

season_levels <- levels(E_fit_season$season)
n_seasons     <- length(season_levels)

E_pos <- E_fit_season %>%
  filter(sum_gaps > 0) %>%
  mutate(log_sum_gaps = log(sum_gaps))

day_seq   <- seq(2, 28, length.out = 100)
dph_seq_c <- day_seq - mean(E_fit_season$day_prior_hatch)

d_freq <- list(
  n_gaps    = E_fit_season$n_gaps,
  dph_c     = E_fit_season$dph_c,
  season_id = as.integer(E_fit_season$season)
)

d_occ <- list(
  has_gap   = E_fit_season$has_gap,
  dph_c     = E_fit_season$dph_c,
  season_id = as.integer(E_fit_season$season)
)

d_dur <- list(
  log_sum_gaps = E_pos$log_sum_gaps,
  dph_c        = E_pos$dph_c,
  season_id    = as.integer(E_pos$season)
)
```

```{r prior-check-seasonal, fig.width=10, fig.height=4, fig.cap="Prior predictive simulations for seasonal models."}
set.seed(123)
n_sim <- 50

par(mfrow = c(1, 3))

# Frequency prior
plot(NULL, xlim = c(28, 2), ylim = c(0, 15),
     xlab = "Days prior hatch", ylab = "Expected n gaps",
     main = "Prior: Frequency")
for(i in 1:n_sim){
  a <- rnorm(1, 1, 0.5); b <- rnorm(1, 0, 0.05)
  lines(day_seq, exp(a + b * dph_seq_c), col = col.alpha("black", 0.3))
}

# Occupancy prior
plot(NULL, xlim = c(28, 2), ylim = c(0, 1),
     xlab = "Days prior hatch", ylab = "P(neglect)",
     main = "Prior: Probability")
for(i in 1:n_sim){
  a <- rnorm(1, 0, 1); b <- rnorm(1, 0, 0.2)
  lines(day_seq, inv_logit(a + b * dph_seq_c), col = col.alpha("black", 0.3))
}

# Duration prior
plot(NULL, xlim = c(28, 2), ylim = c(3, 10),
     xlab = "Days prior hatch", ylab = "log Duration (sec)",
     main = "Prior: Duration (log scale)")
for(i in 1:n_sim){
  a <- rnorm(1, 6, 0.5); b <- rnorm(1, 0, 0.1)
  lines(day_seq, a + b * dph_seq_c, col = col.alpha("black", 0.3))
}

par(mfrow = c(1, 1))
```

```{r fit-seasonal-frequency, include=FALSE, cache=TRUE}
m_season_freq <- quap(
  alist(
    n_gaps ~ dpois(lambda),
    log(lambda) <- a[season_id] + b * dph_c,
    a[season_id] ~ dnorm(1, 0.5),
    b ~ dnorm(0, 0.05)
  ),
  data  = d_freq,
  start = list(a = rep(1, n_seasons), b = 0)
)
```

```{r fit-seasonal-occupancy, include=FALSE, cache=TRUE}
m_season_occ <- quap(
  alist(
    has_gap ~ dbinom(1, p),
    logit(p) <- a[season_id] + b * dph_c,
    a[season_id] ~ dnorm(0, 1),
    b ~ dnorm(0, 0.2)
  ),
  data  = d_occ,
  start = list(a = rep(0, n_seasons), b = 0)
)
```

```{r fit-seasonal-duration, include=FALSE, cache=TRUE}
m_season_dur <- quap(
  alist(
    log_sum_gaps ~ dnorm(mu, sigma),
    mu <- a[season_id] + b * dph_c,
    a[season_id] ~ dnorm(6, 0.5),
    b ~ dnorm(0, 0.1),
    sigma ~ dexp(1)
  ),
  data  = d_dur,
  start = list(a = rep(6, n_seasons), b = 0, sigma = 1)
)
```

```{r posterior-seasonal, include=FALSE}
post_freq <- extract.samples(m_season_freq)
post_occ  <- extract.samples(m_season_occ)
post_dur  <- extract.samples(m_season_dur)

# Frequency predictions
mu_freq <- list(); PI_freq <- list()
for(s in 1:n_seasons){
  l <- link(m_season_freq, data = list(season_id = rep(s, 100), dph_c = dph_seq_c))
  mu_freq[[s]] <- apply(l, 2, mean)
  PI_freq[[s]] <- apply(l, 2, PI, prob = 0.89)
}

# Occupancy predictions
mu_occ <- list(); PI_occ <- list()
for(s in 1:n_seasons){
  l <- link(m_season_occ, data = list(season_id = rep(s, 100), dph_c = dph_seq_c))
  mu_occ[[s]] <- apply(l, 2, mean)
  PI_occ[[s]] <- apply(l, 2, PI, prob = 0.89)
}

# Duration predictions (back-transformed)
mu_dur <- list(); PI_dur <- list()
for(s in 1:n_seasons){
  l <- exp(link(m_season_dur, data = list(season_id = rep(s, 100), dph_c = dph_seq_c)))
  mu_dur[[s]] <- apply(l, 2, mean)
  PI_dur[[s]] <- apply(l, 2, PI, prob = 0.89)
}
```

```{r results-seasonal}
cat("=== FREQUENCY COMPARISONS ===\n")
for(i in 1:(n_seasons - 1)){
  for(j in (i + 1):n_seasons){
    diff <- post_freq$a[,i] - post_freq$a[,j]
    cat(season_levels[i], "vs", season_levels[j],
        "| log diff:", round(mean(diff), 3),
        "| P(S1 > S2):", round(mean(diff > 0), 3), "\n")
  }
}

cat("\n=== OCCUPANCY COMPARISONS ===\n")
for(i in 1:(n_seasons - 1)){
  for(j in (i + 1):n_seasons){
    diff <- post_occ$a[,i] - post_occ$a[,j]
    cat(season_levels[i], "vs", season_levels[j],
        "| logit diff:", round(mean(diff), 3),
        "| P(S1 > S2):", round(mean(diff > 0), 3), "\n")
  }
}

cat("\n=== DURATION COMPARISONS ===\n")
for(i in 1:(n_seasons - 1)){
  for(j in (i + 1):n_seasons){
    diff <- post_dur$a[,i] - post_dur$a[,j]
    cat(season_levels[i], "vs", season_levels[j],
        "| log diff:", round(mean(diff), 3),
        "| P(S1 > S2):", round(mean(diff > 0), 3), "\n")
  }
}

cat("\n=== GLOBAL SLOPES ===\n")
cat("Frequency slope b:", round(mean(post_freq$b), 4),
    " 89% PI:", round(PI(post_freq$b), 4),
    " P(b < 0):", round(mean(post_freq$b < 0), 3), "\n")
cat("Occupancy slope b:", round(mean(post_occ$b), 4),
    " 89% PI:", round(PI(post_occ$b), 4),
    " P(b < 0):", round(mean(post_occ$b < 0), 3), "\n")
cat("Duration slope b: ", round(mean(post_dur$b), 4),
    " 89% PI:", round(PI(post_dur$b), 4),
    " P(b < 0):", round(mean(post_dur$b < 0), 3), "\n")
```

Gap frequency differed substantially across seasons, with 2019 showing
the highest baseline rate of nest exits and 2025 the lowest. The
difference between 2019 and 2025 was the largest observed (log diff:
0.868; P(2019 > 2025) = 1.0), and 2019 was credibly higher than all
other seasons (P > 0.997 in all comparisons). Differences among the
remaining seasons were smaller and less certain, with 2020, 2021, and
2023 showing broadly similar frequency levels (P ranging from 0.77 to
0.97), though all were credibly higher than 2025 (P = 1.0 in all cases).

The probability of any neglect occurring on a given day showed weaker
and less consistent seasonal differences than frequency. No pairwise
comparison was fully credible, with posterior probabilities ranging from
0.31 to 0.939. The clearest signal was between 2020 and 2025 (logit
diff: 0.664; P(2020 > 2025) = 0.939) and between 2021 and 2025 (logit
diff: 0.584; P = 0.890), suggesting 2025 tended to have a lower
probability of neglect, consistent with the frequency results. Seasons
2019, 2020, and 2021 were largely indistinguishable from each other in
terms of neglect probability.

Neglect duration when it occurred showed a different pattern from
frequency. The most notable contrasts involved 2025, which showed
substantially shorter neglect episodes than 2021 (log diff: 1.124;
P(2021 > 2025) = 0.999) and 2023 (log diff: 1.198; P(2023 > 2025) =
0.999). In contrast, 2019 and 2020 showed similar durations to each
other (log diff: −0.020; P = 0.472) and were intermediate between the
longer-duration seasons (2021, 2023) and the shorter-duration season
(2025). Differences between 2019/2020 and 2021/2023 were suggestive but
uncertain (P ranging from 0.064 to 0.088).

Across all three models, the global slope for days prior to hatch was
positive and credible (frequency: β = 0.165, 89% PI [0.148, 0.182];
occupancy: β = 0.212, 89% PI [0.122, 0.301]; duration: β = 0.222, 89%
PI [0.151, 0.292]; P(β < 0) = 0 in all cases), confirming that neglect
— in all its dimensions — consistently declines as hatching approaches,
and that this pattern is shared across all seasons.

```{r plot-seasonal-frequency, fig.width=8, fig.height=5, fig.cap="Expected number of nest exits per day by season across incubation. Lines show posterior mean with 89% PI."}
plot(NULL, xlim = c(28, 2),
     ylim = c(0, max(E_fit_season$n_gaps) * 1.2),
     xlab = "Days prior hatch", ylab = "Number of nest exits")
for(s in 1:n_seasons){
  idx <- E_fit_season$season == season_levels[s]
  points(jitter(E_fit_season$day_prior_hatch[idx], amount = 0.25),
         jitter(E_fit_season$n_gaps[idx], amount = 0.1),
         pch = 16, col = col.alpha(cols_season[s], 0.35), cex = 0.8)
  shade(PI_freq[[s]], day_seq, col = col.alpha(cols_season[s], 0.12))
  lines(day_seq, mu_freq[[s]], lwd = 2.5, col = cols_season[s])
}
legend("topright", legend = season_levels,
       col = cols_season[1:n_seasons], pch = 16, lwd = 2.5, bty = "n")
```

```{r plot-seasonal-occupancy, fig.width=8, fig.height=5, fig.cap="Probability of any neglect per day by season across incubation."}
plot(NULL, xlim = c(28, 2), ylim = c(0, 1),
     xlab = "Days prior hatch", ylab = "P(neglect)")
for(s in 1:n_seasons){
  shade(PI_occ[[s]], day_seq, col = col.alpha(cols_season[s], 0.12))
  lines(day_seq, mu_occ[[s]], lwd = 2.5, col = cols_season[s])
}
legend("topright", legend = season_levels,
       col = cols_season[1:n_seasons], lwd = 2.5, bty = "n")
```

```{r plot-seasonal-duration, fig.width=8, fig.height=5, fig.cap="Neglect duration (seconds, log scale) given neglect occurred, by season across incubation."}
plot(NULL, xlim = c(28, 2), ylim = c(0, 10000),
     xlab = "Days prior hatch", ylab = "Duration of neglect (sec)")
for(s in 1:n_seasons){
  idx <- E_pos$season == season_levels[s]
  points(jitter(E_pos$day_prior_hatch[idx], amount = 0.25),
         E_pos$sum_gaps[idx],
         pch = 16, col = col.alpha(cols_season[s], 0.35), cex = 0.8)
  shade(PI_dur[[s]], day_seq, col = col.alpha(cols_season[s], 0.12))
  lines(day_seq, mu_dur[[s]], lwd = 2.5, col = cols_season[s])
}
legend("topright", legend = season_levels,
       col = cols_season[1:n_seasons], pch = 16, lwd = 2.5, bty = "n")
```